An Effective Model of HL-60 Differentiation

### 2017-01-24 Paper_v2.tex
Working on MM

\section*{Results}
We constructed an effective model of ATRA-induced HL-60 differentiation which described signaling and gene expression events following the addition of ATRA (Fig. \ref{fig:network}).
The model connectivity was developed from literature and the studies presented here (Table ZZ).
We decomposed the ATRA program into three modules;
a signal initiation module that sensed and transformed the ATRA signal into activated cRaf-pS621 and the ATRA-RXR/RAR (Trigger) signals (Fig. \ref{fig:network}A);
a signal integration module that controlled the expression of upstream transcription factors given cRaf-pS621 and activated Trigger signals (Fig. \ref{fig:network}B); and
a phenotype module which encoded the expression of functional differentiation markers from the ATRA-inducible transcription factors (Fig. \ref{fig:network}C).
Each component of these modules was described by a mRNA and protein balance equation.
Additionally, the signal initiation module also described the abundance of activated species e.g., Trigger and cRaf-pS621 whose values were derived from
unactivated Trigger and cRaf protein levels.
Lastly, because the population of HL-60 cells was dividing (at least before ATRA-induced cell cycle arrest), we also considered a dilution term in all balance equations.
The signal initiation module contained nine differential equations, while the signal integration and phenotype modules were collectively encoded by 54 differential equations.
Model parameters were taken literature, or estimated from experimental data taken from literature using heuristic optimization (see materials and methods).

@@

The signal integration and phenotype modules described ATRA-induced gene expression events in wild-type HL-60 cells (Fig. \ref{fig:model-grn-simulations}).
The signal initiation module produced two outputs, activated Trigger and cRaf-pS621 which drove the expression of ATRA-induced transcription factors,
which then in turn activated the phenotypic program. In particular, Trigger, which is a surrogate for the RAR$\alpha$/RXR transcriptional complex,
regulated the expression of the transcription factors CCATT/enhancer binding protein $\alpha$ (C/EBP$\alpha$), PU.1, and EGR1.
In turn, these transcription factors, in combination with cRaf-pS621, regulated the expression of downstream phenotypic markers such as CD38, CD11b or P47Phox.
We assembled the connectivity of the signal integration and phenotypic programs driven by Trigger and cRaf-pS621 from literature (Table ZZ).
We estimated the parameters which appeared in the control laws regulating these programs
from steady-state and dynamic measurements of transcription factor and phenotypic marker expression following the addition of ATRA [REFHERE].
However, the bulk of the remaining model parameters were taken from directly from literature [REFHERE] and were not estimated in this study (see materials and methods).
The model simulations captured the time dependent expression of CD38 and CD11b following the addition ATRA (Fig. \ref{fig:model-grn-simulations}A),
and the steady-state for signal integration and phenotypic markers (Fig. \ref{fig:model-grn-simulations}B).
Taken together, the signal integration and phenotypic simulations were consistent with measurements, thereby validating the assumed molecular connectivity.

@@

\subsubsection*{Model equations.}
We decomposed the ATRA-induced differentiation program into three modules;
a signal initiation module that sensed and transformed the ATRA signal into activated cRaf-pS621 and the ATRA-RXR/RAR (activated Trigger) signals;
a signal integration module that controlled the expression of upstream transcription factors given cRaf-pS621 and activated Trigger signals; and
a phenotype module which encoded the expression of functional differentiation markers from the ATRA-inducible transcription factors.
The output of the signal initiation module was the input to the gene expression model.
For each gene $j=1,2,\dots,\mathcal{G}$, we modeled both the mRNA ($m_{j}$) and protein ($p_{j}$) abundance:
\begin{eqnarray}
	\frac{dm_{j}}{dt} &=& r_{T,j} - \left(\mu+\theta_{m,j}\right)m_{j}+\lambda_{j}\\
	\frac{dp_{j}}{dt} &=& r_{X,j} - \left(\mu+\theta_{p,j}\right)p_{j}
\end{eqnarray}
The terms $r_{T,j}$ and $r_{X,j}$ denote the specific rates of transcription, and translation while
the terms $\theta_{m,j}$ and $\theta_{p,j}$ denote first-order degradation constants for mRNA and protein, respectively.
The specific transcription rate $r_{T,j}$ was modeled as the product of a kinetic term $\bar{r}_{T,j}$ and a control term $u_{j}$ which described how the
abundance of transcription factors, or other regulators influenced the expression of gene $j$.
The kinetic transcription term $\bar{r}_{T,j}$ was modeled as:
\begin{equation}
	\bar{r}_{T,j} = V^{max}_{T}\left(\frac{L_{T,o}}{L_{T,j}}\right)\left(\frac{G_{j}}{K_{T}+G_{j}}\right)
\end{equation}where the maximum gene expression rate $V^{max}_{T}$ was defined as the product of a characteristic transcription rate constant ($k_{T}$)
and the abundance of RNA polymerase ($R_{1}$), $V^{max}_{T} = k_{T}\left(R_{1}\right)$.
The $\left(L_{T,o}/L_{T,j}\right)$ term denotes the ratio of transcription read lengths; $L_{T,o}$ represents a characteristic gene length, while $L_{T,j}$ denotes the length of gene $j$.
Thus, the ratio $\left(L_{T,o}/L_{T,j}\right)$ is a gene specific correction to the characteristic transcription rate $V^{max}_{T}$.
The degradation rate constants were defined as $\theta_{m,j}$ and $\theta_{p,j}$ denote characteristic degradation constants for mRNA and protein, respectively.
Lastly, the $\lambda_{j}$ term denotes the constitutive rate of expression of gene $j$.

@@

The signal initiation, and integration modules required the level of cRaf-pS621 and ATRA-RXR/RAR (activated Trigger) as inputs.
However, our base model described only the abundance of inactive proteins e.g., cRaf or RXR/RAR but not the activated forms.
To address this issue, we estimated pseudo steady state approximations for the abundance of cRaf-pS621 and activated Trigger.
The abundance of activated trigger ($x_{a,1}$) was estimated directly from the RXR/RAR abundance ($x_{u,1}$):
\begin{equation}
	x_{a,1} \sim x_{u,1}\left(\frac{\alpha\times{\texttt{ATRA}}}{1+\alpha\times{ATRA}}\right)
\end{equation}
On the other hand, the abundance of cRaf-pS621 was estimated by making the pseudo steady state approximation on the cRaf-pS621 balance.
The abundance of an activated signaling species $i$ was given by:
\begin{equation}
	\frac{dx_{i}}{dt}  =  r_{+,i}\left(\mathbf{x},\mathbf{k}\right) - \left(\mu+k_{d,i}\right)x_{i}\qquad{i=1,\hdots,\mathcal{M}}\\
\end{equation}
The quantity $x_{i}$ denotes concentration of signaling species $i$, while $\mathcal{R}$ and $\mathcal{M}$ denote the number of signaling reactions and signaling species in the model,
respectively. The term $r_{+,i}\left(\mathbf{x},\mathbf{k}\right)$ denotes the rate of generation of activated species $i$,
while  $\mu$ denotes the specific growth rate, and $k_{d,i}$ denotes the rate constant controlling the non-specific degradation of $x_{i}$.
We have neglected deactivation reactions e.g., phosphatase activities.
We assumed that signaling processes were fast compared to gene expression; this allowed us to approximate the signaling balance as:
\begin{equation}
	x_{i}^{*} \simeq \frac{r_{+,i}\left(\mathbf{x},\mathbf{k}\right)}{\left(\mu+k_{d,i}\right)} \qquad{i=1,\hdots,\mathcal{M}}
\end{equation}
The generation rate was written as the product of a kinetic term ($\bar{r}_{+,i}$) and a control term ($v_{i}$).
The control terms $0\leq v_{j}\leq 1$ depended upon the combination of factors which influenced rate process $j$.
If rate $j$ was influenced by $1,\dots,m$ factors, we modeled this relationship as
$v_{j}=\mathcal{I}_{j}\left(f_{1j}\left(\cdot\right),\hdots,f_{mj}\left(\cdot\right)\right)$
where $0\leq f_{ij}\left(\cdot\right)\leq 1$ denotes a regulatory transfer function quantifying the influence of factor $i$ on rate $j$.
The function $\mathcal{I}_{j}\left(\cdot\right)$ is an integration rule which maps the output of regulatory transfer functions into a control
variable. In this study, we used $\mathcal{I}_{j}\in\left\{min,max\right\}$ and hill transfer functions \cite{pr3010138}.
If a process had no modifying factors, $v_{j}=1$.
The kinetic rate of cRaf-pS621 generation $\bar{r}_{+,cRaf}$ was given as:
\begin{equation}
	\bar{r}_{+,cRaf} = k_{+,cRaf}x_{s}\left(\frac{x_{cRaf}}{K_{+,cRaf}+x_{cRaf}}\right)
\end{equation}where $x_{s}$ denotes the signalsome abundance, and $K_{+,cRaf}$ denotes a saturation constant governing cRaf-pS621 formation.
The formation of cRaf-pS621 was regulated by a single factor, the abundance of MAPK inhibitor, thus $v_{+,cRaf}$ took the form:
\begin{equation}
	v_{+,cRaf} = \left(1-\frac{I}{K_{D}+I}\right)
\end{equation}where $I$ denotes the MAPK inhibitor abundance, and $K_{D}$ the affinity constant for inhibitor $I$.

@@

\noindent\subsubsection*{Estimation of signaling and gene expression model parameters.}
We estimated the $W_{ij}$ parameters, and the parameters in the transfer functions $f_{dj}$ from gene expression data sets.
On the other hand, we estimated $k_{T}, k_{X}, \theta_{m,j}, \theta_{p,j}, R_{1}$ and $R_{2}$ using estimates of transcription and translation rates,
the half-life of a typical mRNA and protein, and a typical value for the copies per cell of RNA polymerase and ribosomes from literature \cite{Milo:2010aa}.
The saturation constants $K_{X}$ and $K_{T}$ were adjusted so that gene expression and translation resulted in gene products on a biologically realistic concentration scale.
Lastly, we calculated the concentration for gene $G_{j}$ by assuming, on average, that a cell had two copies of each gene at any given time.
Thus, the bulk of our gene expression parameters were based directly upon literature values, and were not adjusted during model identification.
The values used for the characteristic transcription/translation parameters, degradation constants and
macromolecular copy number are given in the supplemental materials along with the specific formulas required to calculate all derived constants.

@@

>>>>

\section*{Results}
We constructed an effective model of ATRA-induced HL-60 differentiation which described signaling and gene expression events following the addition of ATRA (Fig. \ref{fig:network}).
The model connectivity was developed from literature and the studies presented here (Table \ref{tbl:TF-network-connectivity}).
We decomposed the ATRA program into three modules;
a signal initiation module that sensed and transformed the ATRA signal into activated cRaf-pS621 and the ATRA-RXR/RAR (Trigger) signals (Fig. \ref{fig:network}A);
a signal integration module that controlled the expression of upstream transcription factors given cRaf-pS621 and activated Trigger signals (Fig. \ref{fig:network}B); and
a phenotype module which encoded the expression of functional differentiation markers from the ATRA-inducible transcription factors (Fig. \ref{fig:network}C).
Each component of these modules was described by a mRNA and protein balance equation.
Additionally, the signal initiation module also described the abundance of activated species e.g., Trigger and cRaf-pS621 whose values were derived from
unactivated Trigger and cRaf protein levels.
Lastly, because the population of HL-60 cells was dividing (at least before ATRA-induced cell cycle arrest), we also considered a dilution term in all balance equations.
The signal initiation module contained nine differential equations, while the signal integration and phenotype modules were collectively encoded by 54 differential equations.
Model parameters were taken literature, or estimated from experimental data taken from literature using heuristic optimization (see materials and methods).

@@

The signal integration and phenotype modules described ATRA-induced gene expression events in wild-type HL-60 cells (Fig. \ref{fig:model-grn-simulations}).
The signal initiation module produced two outputs, activated Trigger and cRaf-pS621 which drove the expression of ATRA-induced transcription factors,
which then in turn activated the phenotypic program. In particular, Trigger, which is a surrogate for the RAR$\alpha$/RXR transcriptional complex,
regulated the expression of the transcription factors CCATT/enhancer binding protein $\alpha$ (C/EBP$\alpha$), PU.1, and EGR1.
In turn, these transcription factors, in combination with cRaf-pS621, regulated the expression of downstream phenotypic markers such as CD38, CD11b or P47Phox.
We assembled the connectivity of the signal integration and phenotypic programs driven by Trigger and cRaf-pS621 from literature (Table \ref{tbl:TF-network-connectivity}).
We estimated the parameters which appeared in the control laws regulating these programs
from steady-state and dynamic measurements of transcription factor and phenotypic marker expression following the addition of ATRA [REFHERE].
However, the bulk of the remaining model parameters were taken from directly from literature \cite{Milo:2010aa} and were not estimated in this study (see materials and methods).
The model simulations captured the time dependent expression of CD38 and CD11b following the addition ATRA (Fig. \ref{fig:model-grn-simulations}A),
and the steady-state for signal integration and phenotypic markers (Fig. \ref{fig:model-grn-simulations}B).
Taken together, the signal integration and phenotypic simulations were consistent with measurements, thereby validating the assumed molecular connectivity.

@@

\subsubsection*{Gene expression model equations.}
We decomposed the ATRA-induced differentiation program into three modules;
a signal initiation module that sensed and transformed the ATRA signal into activated cRaf-pS621 and the ATRA-RXR/RAR (activated Trigger) signals;
a signal integration module that controlled the expression of upstream transcription factors given cRaf-pS621 and activated Trigger signals; and
a phenotype module which encoded the expression of functional differentiation markers from the ATRA-inducible transcription factors.
The output of the signal initiation module was the input to the gene expression model.
For each gene $j=1,2,\dots,\mathcal{G}$, we modeled both the mRNA ($m_{j}$) and protein ($p_{j}$) abundance:
\begin{eqnarray}
	\frac{dm_{j}}{dt} &=& r_{T,j} - \left(\mu+\theta_{m,j}\right)m_{j}+\lambda_{j}\\
	\frac{dp_{j}}{dt} &=& r_{X,j} - \left(\mu+\theta_{p,j}\right)p_{j}
\end{eqnarray}
The terms $r_{T,j}$ and $r_{X,j}$ denote the specific rates of transcription, and translation while
the terms $\theta_{m,j}$ and $\theta_{p,j}$ denote first-order degradation constants for mRNA and protein, respectively.
The specific transcription rate $r_{T,j}$ was modeled as the product of a kinetic term $\bar{r}_{T,j}$ and a control term $u_{j}$ which described how the
abundance of transcription factors, or other regulators influenced the expression of gene $j$.
The kinetic transcription term $\bar{r}_{T,j}$ was modeled as:
\begin{equation}
	\bar{r}_{T,j} = V^{max}_{T}\left(\frac{L_{T,o}}{L_{T,j}}\right)\left(\frac{G_{j}}{K_{T}+G_{j}}\right)
\end{equation}where the maximum gene expression rate $V^{max}_{T}$ was defined as the product of a characteristic transcription rate constant ($k_{T}$)
and the abundance of RNA polymerase ($R_{1}$), $V^{max}_{T} = k_{T}\left(R_{1}\right)$.
The $\left(L_{T,o}/L_{T,j}\right)$ term denotes the ratio of transcription read lengths; $L_{T,o}$ represents a characteristic gene length, while $L_{T,j}$ denotes the length of gene $j$.
Thus, the ratio $\left(L_{T,o}/L_{T,j}\right)$ is a gene specific correction to the characteristic transcription rate $V^{max}_{T}$.
The degradation rate constants were defined as $\theta_{m,j}$ and $\theta_{p,j}$ denote characteristic degradation constants for mRNA and protein, respectively.
Lastly, the $\lambda_{j}$ term denotes the constitutive rate of expression of gene $j$.

@@

\subsubsection*{Signaling model equations.}
The signal initiation, and integration modules required the level of cRaf-pS621 and ATRA-RXR/RAR (activated Trigger) as inputs.
However, our base model described only the abundance of inactive proteins e.g., cRaf or RXR/RAR but not the activated forms.
To address this issue, we estimated pseudo steady state approximations for the abundance of cRaf-pS621 and activated Trigger.
The abundance of activated trigger ($x_{a,1}$) was estimated directly from the RXR/RAR abundance ($x_{u,1}$):
\begin{equation}
	x_{a,1} \sim x_{u,1}\left(\frac{\alpha\cdot{\mathrm{ATRA}}}{1+\alpha\cdot\mathrm{ATRA}}\right)
\end{equation}
where $\alpha$ denotes a gain parameter; $\alpha = 0.0$ if ATRA is less than a threshold, and $\alpha = 0.1$ if ATRA is greater than the differentiation threshold.
The abundance of cRaf-pS621 was estimated by making the pseudo steady state approximation on the cRaf-pS621 balance.
The abundance of an activated signaling species $i$ was given by:
\begin{equation}
	\frac{dx_{i}}{dt}  =  r_{+,i}\left(\mathbf{x},\mathbf{k}\right) - \left(\mu+k_{d,i}\right)x_{i}\qquad{i=1,\hdots,\mathcal{M}}\\
\end{equation}
The quantity $x_{i}$ denotes concentration of signaling species $i$, while $\mathcal{R}$ and $\mathcal{M}$ denote the number of signaling reactions and signaling species in the model,
respectively. The term $r_{+,i}\left(\mathbf{x},\mathbf{k}\right)$ denotes the rate of generation of activated species $i$,
while $\mu$ denotes the specific growth rate, and $k_{d,i}$ denotes the rate constant controlling the non-specific degradation of $x_{i}$.
We neglected deactivation reactions e.g., phosphatase activities.
We assumed that signaling processes were fast compared to gene expression; this allowed us to approximate the signaling balance as:
\begin{equation}
	x_{i}^{*} \simeq \frac{r_{+,i}\left(\mathbf{x},\mathbf{k}\right)}{\left(\mu+k_{d,i}\right)} \qquad{i=1,\hdots,\mathcal{M}}
\end{equation}
The generation rate was written as the product of a kinetic term ($\bar{r}_{+,i}$) and a control term ($v_{i}$).
The control terms $0\leq v_{j}\leq 1$ depended upon the combination of factors which influenced rate process $j$.
If rate $j$ was influenced by $1,\dots,m$ factors, we modeled this relationship as
$v_{j}=\mathcal{I}_{j}\left(f_{1j}\left(\cdot\right),\hdots,f_{mj}\left(\cdot\right)\right)$
where $0\leq f_{ij}\left(\cdot\right)\leq 1$ denotes a regulatory transfer function quantifying the influence of factor $i$ on rate $j$.
The function $\mathcal{I}_{j}\left(\cdot\right)$ is an integration rule which maps the output of regulatory transfer functions into a control
variable. In this study, we used $\mathcal{I}_{j}\in\left\{min,max\right\}$ and hill transfer functions \cite{pr3010138}.
If a process had no modifying factors, $v_{j}=1$.
The kinetic rate of cRaf-pS621 generation $\bar{r}_{+,cRaf}$ was modeled as:
\begin{equation}
	\bar{r}_{+,cRaf} = k_{+,cRaf}x_{s}\left(\frac{x_{cRaf}}{K_{+,cRaf}+x_{cRaf}}\right)
\end{equation}where $x_{s}$ denotes the signalsome abundance, and $K_{+,cRaf}$ denotes a saturation constant governing cRaf-pS621 formation.
The formation of cRaf-pS621 was regulated by only a single factor, the abundance of MAPK inhibitor, thus $v_{+,cRaf}$ took the form:
\begin{equation}
	v_{+,cRaf} = \left(1-\frac{\mathrm{I}}{K_{D}+\mathrm{I}}\right)
\end{equation}where $\mathrm{I}$ denotes the abundance of the MAPK inhibitor, and $K_{D}$ denotes the inhibitor affinity.

@@

\noindent\subsubsection*{Estimation of model parameters.}
We estimated parameters appearing in the mRNA and protein balances, and the abundance of polymerases and ribosomes, from estimates of transcription and translation rates,
the half-life of a typical mRNA and protein, and typical values for the copies per cell of RNA polymerase and ribosomes from literature (Table ZZ).
For the remaining parameters, e.g., the $W_{ij}$ appearing in the control laws,
or parameters appearing in the transfer functions $f_{dj}$, were estimated from the gene expression and signaling data sets discussed here.
The saturation constants $K_{X}$ and $K_{T}$ were adjusted so that gene expression and translation resulted in gene products on a biologically realistic concentration scale.
Lastly, we calculated the concentration for gene $G_{j}$ by assuming, on average, that a cell had two copies of each gene at any given time.
Thus, the bulk of our gene expression parameters were based directly upon literature values, and were not adjusted during model identification.
The values used for the characteristic transcription/translation parameters, degradation constants and
macromolecular copy number are given in the supplemental materials along with the specific formulas required to calculate all derived constants.

@@

\begin{center}
\renewcommand{\arraystretch}{0.8}
\tablefirsthead{
\hline
\rowcolor[gray]{0.8}\textbf{Actor}  & \textbf{Effect} & \textbf{Target} & \textbf{Cite}  \\
\hline}
\topcaption{Myelomonocytic transcription factor connectivity used in the signal integration and phenotype modules.}\label{tbl:TF-network-connectivity}
\begin{scriptsize}
\begin{supertabular}{c|c|l|l}
\hline
RAR$\alpha$ &  + & RAR$\alpha$ & \cite{Rishi1996} \\
 	&  + & PU.1 & \cite{Mueller2006}\\
 	&  + & C/EBP$\alpha$ & \cite{Friedman2007}\\
 	&  + & IRF-1 & \cite{Luo2006}\\
 	&  - & Oct4 & \cite{Sylvester1994}\\
 	&  + & CD38 & \cite{Drach1994}\\
 	&  + & p21 & \cite{Liu1996}\\
 	&  + & AhR & \cite{Bunaciu2013}\\
 	&  + & EGR1 & \cite{Balmer2002}\\
\hline
PPAR$\gamma$ &  + & C/EBP$\alpha$ & \cite{Rosen2002}\\
 &  + & IRF-1 & \cite{Varley2009}\\
 &  + & Oct1 & \cite{Bruemmer2003}\\
 &  - & AP-1 & \cite{Delerive1999}\\
 &  - & E2F & \cite{Altiok1997}\\
 &  - & EGR1 & \cite{Fei2011}\\
 &  + & CD38 & \cite{Song2012}\\
 &  + & CD14 & \cite{Szanto2005}\\
 &  + & p21 & \cite{Han2004}\\
 &  - & p47phox & \cite{Von-Knethen2002}\\
\hline
PU.1 &  - & PPAR$\gamma$ & \cite{Dispirito2013}\\
&  + & PU.1 & \cite{Chen1995}\\
&  + & AP-1 & \cite{Steidl2006}\\
&  + & EGR1 & \cite{Laslo2006}\\
&  + & CD11b & \cite{Pahl1993}\\
&  + & p21 & \cite{Yuki2013}\\
&  + & p47phox & \cite{Li1999}\\
\hline
C/EBP$\alpha$ &  + & PPAR$\gamma$ & \cite{Rosen2002}\\
 &  + & PU.1 & \cite{Dahl2003}\\
 &  + & C/EBP$\alpha$ & \cite{Timchenko1995}\\
 &  + & Gfi-1 & \cite{Lidonnici2010}\\
 &  - & E2F & \cite{DAlo2003}\\
 &  + & CD14 & \cite{Pan1999}\\
 &  + & p21 & \cite{Harris2001}\\
\hline
IRF-1 &  + & CD38 & \cite{Bauvois1999}\\
 &  + & p21 & \cite{Passioura2005}\\
 &  - & PU.1 & \cite{Dahl2007}\\
 &  - & C/EBP$\alpha$ & \cite{Duan2003}\\
 &  - & E2F & \cite{Duan2003}\\
 &  - & EGR1 & \cite{Laslo2006}\\
 &  - & p21 & \cite{Duan2003}\\
\hline
Oct1 &  + & PU.1 & \cite{Chen1996}\\
\hline
AP-1 &  - & PPAR$\gamma$  & \cite{Delerive1999}\\
&  + & PU.1 &  \cite{Behre1999}\\
&  + & p21 &  \cite{Kardassis1999}\\
\hline
E2F &  + & E2F & \cite{Johnson1994}\\
\hline
EGR1 &  + & PPAR$\gamma$ & \cite{Fu2002}\\
&  - & Gfi-1 & \cite{Mak2011}\\
&  + & CD14 & \cite{Chen2004}\\
\hline
AhR &  + & AP-1 & \cite{Suh2002}\\
&  + & IRF-1 & \cite{Shen2011}\\
& - & Oct4 & \cite{Bunaciu2011}\\
& - & PU.1 & \\

\hline
\end{supertabular}
\end{scriptsize}
\end{center}
